Global search algorithm with local tuning for solving mixed-integer global optimization problems \thanks{This study was supported by the .}

### 2021-04-08 2021/MOTOR_BarkalovLebedev/MOTOR2021.tex
MOTOR правки

\begin{document}
%
\title{An Approach for Simultaneous Finding of Multiple Effective Decisions in Multi-objective Optimization Problems
%with non-convex constraints 
\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no.0729-2020-0055, and by the Research and Education Mathematical Center, project no. 075-02-2020-1483/1.}}
%
\titlerunning{Finding of Multiple Effective Decisions in MOO Problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Victor Gergel\orcidID{0000-0002-4013-2329} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeniy Kozinov\orcidID{0000-0001-6776-0096} 
}
%
\authorrunning{K. Barkalov, V. Gergel, V. Grishagin, E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia 
\email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru,\{gergel,vagris\}@unn.ru}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This paper considers computationally intensive multi-objective optimization problems which require computing multiple Pareto-optimal decisions. It is also assumed that efficiency criteria may be multiextremal, and the cost of calculating function values may be quite high. The proposed approach is based on the reduction of multi-objective optimization problems to one-dimensional global optimization problems that can be solved using efficient information-statistical algorithms of global search. One key innovation of the developed approach consists in the possibility of solving several global optimization problems simultaneously, which allows multiple Pareto-optimal decisions to be obtained. Besides, such approach provides for reuse of the computed search information, which considerably reduces computational effort for solving multi-objective optimization problems. Computational experiments confirm the potential of the proposed approach.

@@

In the most general form, the MCO problem can be formulated as follows
\begin{equation}
\label{eq:01}
f(y) = (f_1(y), f_2(y), \dots, f_s(y)) \to \min,  y\in D,
\end{equation}
where $f(y) = (f_1(y), f_2(y), \dots, f_s(y))$ are objective functions (efficiency criteria), $y = (y_1, y_2, \dots , y_N)$ is the vector of varied parameters, and $N$ is the dimensionality of the multi-objective optimization problem to be solved. The set of possible parameter values (search domain) $D$ is usually an $N$-dimensional hyperinterval
\begin{equation}
\label{eq:02}
D  = \{ y\in R^N: a_i \leq y_i \leq b_i, 1 \leq i \leq N \}
\end{equation}
for given boundary vectors $a$ and $b$.

@@

\subsection{Scalarization of multiple objective functions.} 
In the most general form, the global optimization problem generated by the scalarization of multiple objective functions of the MOO problem can be represented in the form
\begin{equation}
\label{eq:04}
\min_{y \in D} {\varphi(y)} = \min_{y \in D} F(\lambda, y),
\end{equation}
where $F$ is a scalar multiextremal function generated as a result of scalarization of objective functions $f_i$, $1 \leq i \leq s$, $\lambda$ is the vector of parameters of the applied convolution of functions, and $D$ is the search domain from (\ref{eq:01}). In the proposed approach, we use for scalarization the compromise programming method \cite{c2,c11}, where solution of the MOO problem consists in finding the efficient decision corresponding most closely to the values of the objective functions of the specified reference decision $y^0 \in D$. In this case, a possible statement of scalar optimization problem can have the form 
\begin{equation}
\label{eq:05}
\min_{y \in D} F(\lambda,y)=\min_{y \in D} \left\{ \frac{1}{s} \sum_{i=1}^s {\lambda_i (f_i (y)-f_i (y^0 ))^2} \right\},
\end{equation}
where $F(\lambda,y)$ is the standard deviation of the values of the objective functions $f_i$, $1 \leq i \leq s$, for the decision $y \in D$ and for the specified reference decision $y^0 \in D$, while the coefficients $\lambda_i$, $0 \leq \lambda_i \leq 1$, $1 \leq i \leq s$, are measures of the significance of the approximation accuracy for each variable $y_i$, $1 \leq i \leq N$, separately. Without loss of generality, we can assume that the area of possible values of the coefficients $\lambda$ is a set 
\begin{equation}
\label{eq:06}
\lambda=(\lambda_1,\lambda_2, \dots, \lambda_s)\in \Lambda \subset R^s: \sum_{i=1}^{s}\lambda_i=1, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

@@

\subsection{Dimensionality reduction.} 
As already mentioned, problem (\ref{eq:04}) is a multidimensional global optimization problem. Problems of this kind are computationally complex and are known to be subject to the ``curse of dimensionality'' -- computational complexity increases exponentially with increasing dimensionality of the optimization problem being solved \cite{c13,c14,c15,c16,c17,c18,c19,c20}. Nevertheless, the computational complexity of global optimization algorithms can be significantly reduced by dimensionality reduction based on the use of a Peano space-filling curve (or evolvent) $y(x)$ that uniquely and continuously maps the segment $[0,1]$ on an $N$-dimensional hypercube $D$ -- see, for example, \cite{c15,c21}. As a result of such reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional problems
\begin{equation}
\label{eq:09}
\min_{x \in [0,1]} \varphi (y(x))= \min_{y \in D} \varphi(y).
\end{equation}

@@

\subsection{Solving the one-dimensional reduced optimization problem.} 
Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multi-extremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

@@

In this Section, the proposed approach for finding $p$, $p>1$ Pareto-optimal decisions of the MOO problem is presented. This problem is formulated as the problem of solving the set $\Phi_p (y)$, $p>1$ of global optimization problems (see \ref{subsec:41}). To solve this set of optimization problems, two efficient computational schemes are proposed. The first scheme is based on the traditional sequential procedure for solving multiple problems, however, to solve each subsequent optimization problem, all the search information obtained during previous calculations is taken into account. This accumulated search information can significantly reduce the number of global search iterations required to solve the next optimization problems of the set $\Phi_p (y)$ (see \ref{subsec:42} and Chapter \ref{sec:05}). In the second scheme, all the problems of the set $\Phi_p (y)$ are solved simultaneously, but the search information obtained when solving each problem is also used for solving all the other problems of the set $\Phi_p (y)$ (see \ref{subsec:43}). This useful exchange of search information also significantly reduces the amount of required calculations for solving the problems of the set $\Phi_p (y)$.

\subsection{The need to solve multiple global optimization problems.} 
\label{subsec:41}

@@

\subsection{Step-by-step solution of a set of global optimization problems.} 
\label{subsec:42}

@@

\subsection{Simultaneous solution of a set of global optimization problems.} 
\label{subsec:43}

@@

In the third series of numerical experiments, 10 bi-criteria five-dimensional MOO problems were solved, i.e. $N = 5$, $s = 2$, $m=2$. The objective functions of the MOO problems were determined, as before, using the GKLS generator \cite{c43}. When solving the problems, the accuracy $\varepsilon=0.05$ and the reliability $r=5.6$ were used. The results of numerical experiments are presented in Table \ref{tab:4}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{fig_t5}
  \caption{The average number of iterations for solving bi-criteria five-dimensional MOO problems}
  \label{fig:1}
\end{figure}

>>>>

\begin{document}
%
\title{An Approach for Simultaneous Finding of Multiple Efficient Decisions in Multi-objective Optimization Problems
%with non-convex constraints 
\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no.0729-2020-0055, and by the Research and Education Mathematical Center, project no. 075-02-2020-1483/1.}}
%
\titlerunning{Finding of Multiple Effective Decisions in MOO Problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Victor Gergel\orcidID{0000-0002-4013-2329} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeniy Kozinov\orcidID{0000-0001-6776-0096} 
}
%
\authorrunning{K. Barkalov, V. Gergel, V. Grishagin, E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia 
\email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru,\{gergel,vagris\}@unn.ru}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This paper considers computationally intensive multi-objective optimization problems which require computing multiple Pareto-optimal decisions. It is also assumed that efficiency criteria may be multiextremal, and the cost of calculating function values may be quite high. The proposed approach is based on the reduction of multi-objective optimization problems to one-dimensional global optimization problems that can be solved using efficient information-statistical algorithms of global search. One key innovation of the developed approach consists in the possibility of solving several global optimization problems simultaneously, which allows multiple Pareto-optimal decisions to be obtained. Besides, such approach provides for reuse of the computed search information, which considerably reduces computational effort for solving multi-objective optimization problems. Computational experiments confirm the potential of the proposed approach.

@@

In the most general form, the MOO problem can be formulated as follows
\begin{equation}
\label{eq:01}
f(y) = (f_1(y), f_2(y), \dots, f_s(y)) \to \min,  y\in D,
\end{equation}
where $f(y) = (f_1(y), f_2(y), \dots, f_s(y))$ are objective functions (efficiency criteria), $y = (y_1, y_2, \dots , y_N)$ is the vector of varied parameters, and $N$ is the dimensionality of the multi-objective optimization problem to be solved. The set of possible parameter values (search domain) $D$ is usually an $N$-dimensional hyperinterval
\begin{equation}
\label{eq:02}
D  = \{ y\in R^N: a_i \leq y_i \leq b_i, 1 \leq i \leq N \}
\end{equation}
for given boundary vectors $a$ and $b$.

@@

\subsection{Scalarization of multiple objective functions} 
In the most general form, the global optimization problem generated by the scalarization of multiple objective functions of the MOO problem can be represented in the form
\begin{equation}
\label{eq:04}
\min_{y \in D} {\varphi(y)} = \min_{y \in D} F(\lambda, y),
\end{equation}
where $F$ is a scalar multiextremal function generated as a result of scalarization of objective functions $f_i$, $1 \leq i \leq s$, $\lambda$ is the vector of parameters of the applied convolution of functions, and $D$ is the search domain from (\ref{eq:01}). In the proposed approach, we use for scalarization the compromise programming method \cite{c2,c11}, where solution of the MOO problem consists in finding the efficient decision corresponding most closely to the values of the objective functions of the specified reference decision $y^0 \in D$. In this case, a possible statement of scalar optimization problem can have the form 
\begin{equation}
\label{eq:05}
\min_{y \in D} F(\lambda,y)=\min_{y \in D} \left\{ \frac{1}{s} \sum_{i=1}^s {\lambda_i (f_i (y)-f_i (y^0 ))^2} \right\},
\end{equation}
where $F(\lambda,y)$ is the standard deviation of the values of the objective functions $f_i$, $1 \leq i \leq s$, for the decision $y \in D$ and for the specified reference decision $y^0 \in D$, while the coefficients $\lambda_i$, $0 \leq \lambda_i \leq 1$, $1 \leq i \leq s$, are measures of the significance of the approximation accuracy for each variable $y_i$, $1 \leq i \leq N$, separately. Without loss of generality, we can assume that the area of possible values of the coefficients $\lambda$ is a set 
\begin{equation}
\label{eq:06}
\lambda=(\lambda_1,\lambda_2, \dots, \lambda_s)\in \Lambda \subset R^s: \sum_{i=1}^{s}\lambda_i=1, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

@@

\subsection{Dimensionality reduction} 
As already mentioned, problem (\ref{eq:04}) is a multidimensional global optimization problem. Problems of this kind are computationally complex and are known to be subject to the ``curse of dimensionality'' -- computational complexity increases exponentially with increasing dimensionality of the optimization problem being solved \cite{c13,c14,c15,c16,c17,c18,c19,c20}. Nevertheless, the computational complexity of global optimization algorithms can be significantly reduced by dimensionality reduction based on the use of a Peano space-filling curve (or evolvent) $y(x)$ that uniquely and continuously maps the segment $[0,1]$ on an $N$-dimensional hypercube $D$ -- see, for example, \cite{c15,c21}. As a result of such reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional problems
\begin{equation}
\label{eq:09}
\min_{x \in [0,1]} \varphi (y(x))= \min_{y \in D} \varphi(y).
\end{equation}

@@

\subsection{Solving the one-dimensional reduced optimization problem} 
Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multi-extremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

@@

In this Section, the proposed approach for finding $p$, $p>1$ Pareto-optimal decisions of the MOO problem is presented. This problem is formulated as the problem of solving the set $\Phi_p (y)$, $p>1$ of global optimization problems (see \ref{subsec:41}). To solve this set of optimization problems, two efficient computational schemes are proposed. The first scheme is based on the traditional sequential procedure for solving multiple problems, however, to solve each subsequent optimization problem, all the search information obtained during previous calculations is taken into account. This accumulated search information can significantly reduce the number of global search iterations required to solve the next optimization problems of the set $\Phi_p (y)$ (see \ref{subsec:42} and Section \ref{sec:05}). In the second scheme, all the problems of the set $\Phi_p (y)$ are solved simultaneously, but the search information obtained when solving each problem is also used for solving all the other problems of the set $\Phi_p (y)$ (see \ref{subsec:43}). This useful exchange of search information also significantly reduces the amount of required calculations for solving the problems of the set $\Phi_p (y)$.

\subsection{The need to solve multiple global optimization problems} 
\label{subsec:41}

@@

\subsection{Step-by-step solution of a set of global optimization problems} 
\label{subsec:42}

@@

\subsection{Simultaneous solution of a set of global optimization problems} 
\label{subsec:43}

@@

In the third series of numerical experiments, 10 bi-criteria five-dimensional MOO problems were solved, i.e. $N = 5$, $s = 2$, $m=2$. The objective functions of the MOO problems were determined, as before, using the GKLS generator \cite{c43}. When solving the problems, the accuracy $\varepsilon=0.05$ and the reliability $r=5.6$ were used. The results of numerical experiments are presented in Table \ref{tab:4} and fig. \ref{fig:2}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{fig_t5}
  \caption{The average number of iterations for solving bi-criteria five-dimensional MOO problems}
  \label{fig:2}
\end{figure}
